---
title: "ronaVote"
subtitle: "Ongoing COVID and the 2020 USA Election: Time Trends and Magnitudes of Correlation among the Cross-section of Counties"
author: "Author: Steven Bhardwaj"
date: "2021-08-04"
output:
  html_document:
    theme:
      bootswatch: darkly
    code_folding: hide
---

<!--  
    body { 
      font-family: "Courier New";
    } 
-->

<style type="text/css">
*{
  margin: 0;
  padding: 0;
}
@font-face {
  font-family: "OracleBoneFont_Subset";
  src: url("../ebas927_Subset_PanOnly.ttf") format("truetype");
}

h1.title {
  padding-top: 2em;
  text-align: center;
}
h3.subtitle {
  text-align: center;
}
h4.author {
  font-size: 1em;
  text-align: center;
}
h4.date {
  font-size: 1em;
  text-align: center;
}

#homeLink {
    font-size: 1.5em;
    color: #CCCCCC !important;
    opacity: 1;
    width: 5em;
  	height: 1.2em;
  	
    position: absolute;
    top: 3%;
    left: 50%;
    transform: translate(-50%, 0%);

    
}
</style>

<div id="homeLink"><a href="../index.html"><p align="center">
bhrdwj.net<br>潘
</p></a></div>

## Abstract
In Spring and Summer of 2020, predominantly politically Democratic cities like NYC suffered the worst of the first COVID wave. But after the Fall and the election passed, the winter saw much worse COVID in many Republican-majority areas.

The maximum correlation of Democratic areas and severity of COVID occurred the second week of August 2020, at 0.00008 COVID-cases per percentage point of vote margin. (8 per 100,000) The Democratic and Republican areas had balanced cumulative COVID briefly in the second week of November, at the moment of the election. Then, the maximum correlation of Republican areas with cumulative COVID cases occured in the second or third week of January, at .00017 COVID-cases per percentage point of vote margin. (17 per 100,000)

## Code
```{r, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
library(rmarkdown)
library(data.table)
library(broom)
library(tidyverse)
library(ggdark)
```

#### Import and clean data
```{r}
# Import data
  # Remove rows where FIPS data is missing.
pathRona <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
pathVote <- "https://raw.githubusercontent.com/bhrdj/ronaVote/main/data/vote/Harvard/countypres_2000-2020.csv"
pathPop <- "https://raw.githubusercontent.com/bhrdj/ronaVote/main/data/countyData/USDA_PopulationData/PopEst2019.csv"

ronaDays <- fread(pathRona, select = 
                grep("fips|date|cases",
                names(fread(pathRona, nrow = 0L)))) %>%
            rename(casesTot = cases)
popu <- fread(pathPop, select = 
                grep("fips|state|area_name|pop2019",
                names(fread(pathPop, nrow = 0L))))
vote <- fread(pathVote, select = 
                grep("year|party|mode|candidate|county_name|county_fips|candidatevotes|totalvotes",
                names(fread(pathVote, nrow = 0L)))) %>%
            rename(fips = county_fips)

rawData <- list(ronaDays=ronaDays, popu=popu, vote=vote) %>%
    map(function(x) { filter(x, !is.na(fips)) })
rm(pathRona, pathVote, pathPop, ronaDays, popu, vote)

# Organize and clean tall (melted) COVID data:
  # Assign a number to each week starting with 2019-12-29.
  # Remove excess rows within weeks.
  # Make crosswalks to text-label versions of the numeric dates and FIPS.

zeroMonday <- as.IDate("2019-12-29")
zeroMondayInt <- as.numeric(zeroMonday)

ronaTall <- rawData[["ronaDays"]] %>%
    mutate(weekNum = (as.numeric(date) - as.numeric(zeroMondayInt)) %/% 7) %>%
    distinct(fips, weekNum, .keep_all = TRUE) %>%
    mutate(weekDate = as.IDate(weekNum * 7, origin=zeroMonday)) %>%
    mutate(week_DateT = paste("T", gsub("-", "_", weekDate), sep="")) %>%
    mutate(fipsText = paste("x", fips, sep = ""))

weekDates <- distinct(ronaTall, week_DateT, weekDate)
fips_fipsText <- distinct(ronaTall, fips, fipsText, .keep_all = FALSE)
```

#### Tidy data
```{r}
# Tidy election data:
    # Widen -> 
        # a column for each major party,
        # a row for each FIPS county.
    # Sum votes for FIPS areas with multiple subareas.
    # Calculate vote-margins from vote-counts.

vote_wide <- rawData[["vote"]] %>%
    filter(year == 2020) %>%
    select(-c(year)) %>%
    filter((candidate != "JO JORGENSEN") & (candidate != "OTHER")) %>% 
    pivot_wider(names_from = party, 
                values_from = candidatevotes,
                id_cols = -c(party, candidatevotes, candidate))  %>%
    group_by(fips) %>%
    summarize(DEMOCRAT = sum(DEMOCRAT),
              REPUBLICAN = sum(REPUBLICAN),
              county_name = unique(county_name), 
              totalvotes = unique(totalvotes)) %>%
    mutate(DJT_Margin = (REPUBLICAN - DEMOCRAT) / (REPUBLICAN + DEMOCRAT))

# Tidy COVID data:
    # Widen -> 
        # a column for each week
        # a row for each FIPS county.
    # Join with population and election data.
    # Impute zero-cases for (earlier) weeks where some counties didn't have any cases yet.
ronaSectionsTot <- ronaTall  %>%
    pivot_wider(id_cols = fipsText,
                names_from = week_DateT, 
                values_from = casesTot) %>%
    mutate(across(starts_with("T"), ~replace_na(., 0))) %>%
    left_join(fips_fipsText, by = "fipsText") %>%
    column_to_rownames(var = "fipsText") %>%
    left_join(rawData[["popu"]], by="fips") %>%
    left_join(vote_wide, by="fips")  %>%
    filter(!is.na(pop2019))

# Calculate a column for COVID-cases-per-capita.
rona <- ronaSectionsTot %>%
    mutate(across(starts_with("T"), function(x) {x/pop2019} ))

# Remove unused variables
rm(fips_fipsText, rawData, ronaTall, ronaSectionsTot)
```

#### Regress
```{r}
# Regress COVID cases-per-capita on election data
    # One regression for each week
    # Each regression across the cross-section of FIPS counties
lm_allT <-  map(select(rona, starts_with("T")), 
                       function(yvar) {
                         lm(yvar ~ DJT_Margin, rona)
                         })

# Tidy regression coefficients and other statistical output.
lm_outT <- map(lm_allT, 
               function(an_lm) {
                   c( tidy(an_lm)$estimate[2], 
                      glance(an_lm)$r.squared,
                      glance(an_lm)$p.value )
               })

lm_dfT <- as.data.frame(do.call(rbind, lm_outT))
colnames(lm_dfT) <- c("Correlation_MoreTrumpMargin_MoreCovid", "RSquared", "PValue")
lm_dfT2 <- lm_dfT %>% 
    rownames_to_column(var = "week_DateT") %>%
    left_join(weekDates, by = "week_DateT") %>%
    filter(weekDate > as.Date("2020-04-12"))
```

## Plot
```{r echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
title_ = "Magnitude of Correlation Over Time"
subtitle_ = "(Between Percap COVID and 2020 Vote Margin Across USA Counties)"
caption_ = paste(
  "Each data point is a regression coefficient,\n",
  "at a particular week, across all USA FIPS counties, between:\n",
  "per-capita cumulative COVID cases, and\n",
  "the 2019 vote margin.\n\n",
  "A coefficient of +0.01 means that:\n",
  "an increase of +.01 in vote margin (+Republican) would correlate with\n",
  "an increase of +1 COVID case per 10,000 population.")
```

```{r fig.align='center', warning=FALSE, message=FALSE}
TrumpMargin_vs_CovidPerCap <- lm_dfT2 %>% 
    ggplot(aes(x=weekDate, 
               y=Correlation_MoreTrumpMargin_MoreCovid)) +
      labs(title=title_, subtitle=subtitle_, caption=caption_) +
      xlab("Date") + ylab("Regression Coefficient") +
      scale_y_continuous(limits=c(-.01, .02)) +
      geom_hline(yintercept = 0, linetype = "solid", color = "white", size = .5) +
      geom_point(aes(color=Correlation_MoreTrumpMargin_MoreCovid)) +
      scale_colour_gradient(low="#0000ff", high="#ff0000", guide=FALSE)

TrumpMargin_vs_CovidPerCap + dark_theme_gray()
```

<br>

## Description
Each plotted point on the graph is a cross-section regression coefficient, taken at a point in time. It measures the correlation at that time across the cross-section of all USA FIPS counties, and this is done every week.

The coefficient measures correlation between:
 * per-capita cumulative COVID cases, and
 * the 2019 vote margin.

Of course, the 2019 vote margin doesn't change every week, only the cumulative count of COVID cases can increase. And, the cumulative counts can never decrease, by definition! But, if there is a trend, that counties that voted more-Republican are suffering more COVID cases, then the coefficient will increase with that correlated trend. 

Note that the coefficient was negative in early 2020, when the Democratic counties were experiencing greater vulnerability to the virus.

So, a coefficient of +0.01 means that an increase of +.01 in vote margin (+ defined as + Republican), would correlate with an increase of +1 COVID case per 10,000 population. This "increase" is not happening over time, it is a way of making an incremental comparison between different counties in the cross-section.

## Interpreting the Magnitude of Coefficients
A coefficient of .01 naively implies that an change of 1 in the vote margin correlates with an increase in .01 cases per population. But, the vote margin is only relevant in smaller increments, so let's divide by 100.

Then, we can say that a correlation of .01 implies that an increase of .01 in vote margin correlates with one additional COVID case per 10,000 population.

Consider an example county with 10,000 Americans, including 5,000 voters, and presume a coefficient of +.01, where Republican votes correlate with more COVID cases.

If the county were to hypothetically move along this margin such that there were 25 more Republican voters and 25 fewer Democrat voters, then that would increase the margin by +.01, corresponding to one additional case in the county.

## Correlation, Causality, and References
I think I should leave it to epidemiologists to publicly cast judgement on how to make the correct inference from this correlation.

This observation [has](https://www.nytimes.com/interactive/2021/04/17/us/vaccine-hesitancy-politics.html) been [reported](https://www.npr.org/sections/health-shots/2020/11/06/930897912/many-places-hard-hit-by-covid-19-leaned-more-toward-trump-in-2020-than-2016)  since even [before](https://www.washingtonpost.com/politics/2021/07/27/rate-new-infections-is-about-twice-high-red-counties-blue-counties/) the election, but maybe a fresh take might be relevant.





<!--
# if .005 of voters switch their vote, that's a +.01 margin
# +.01 margin corresponds to a total of 155,146,377 * .005 = 775,732 votes
# +.0001 cases per population corresponds to 322,793,363 * +.0001 = 32,279 cases
# 775,732 votes divided by 32,279 cases equals a marginal rate of about 25 votes per case
-->

<!-- 
# CODE AND INFO FOR LATER ------------------------------------------------------

# library(shiny)
# library(bslib)
# library(thematic)

# pathRona <- "./data/rona/NYT/us-counties_panel_2021-08-02.csv"
# pathVote <- "./data/vote/Harvard/countypres_2000-2020.csv"
# pathPop <- "./data/countyData/USDA_PopulationData/PopEst2019.csv"

# pathRona <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
# pathVote <- "https://raw.githubusercontent.com/bhrdj/ronaVote/main/data/vote/Harvard/countypres_2000-2020.csv"
# pathPop <- "https://raw.githubusercontent.com/bhrdj/ronaVote/main/data/countyData/USDA_PopulationData/PopEst2019.csv"

# library(rsconnect)
# rsconnect::deployApp('~/git/RmdSandbox/')
# https://bhrdj.shinyapps.io/rmdsandbox/

# https://www.shinyapps.io/admin/#/dashboard

# RUN THESE TO CLEAR LOCAL DATA BUILDING UP IN R WORKSPACE
# rm(list = ls())
# .rs.restartR()


# # PLOT -------------------------------------------------------------------------
# thematic_shiny()
# 
# # Define UI for application that draws a histogram
# ui <- fluidPage(
#     # theme = shinytheme("slate"),
#     theme = bs_theme(version = 4, bootswatch = "darkly"),
#     titlePanel("Trend in Correlation Over Time"),
#         # Show a plot of the generated distribution
#         mainPanel(
#            plotOutput("distPlot1")
#         )
# )
# 
# # Define server logic required to draw a histogram
# server <- function(input, output) {
#     output$distPlot1 <- renderPlot({
#         lm_dfT2 %>% ggplot(aes(x=weekDate, y=Correlation_MoreTrumpMargin_MoreCovid)) + geom_point()
#         })
# }
# 
# # Run the application 
# shinyApp(ui = ui, server = server)




-->