---
title: "ronaVote Appendix: Within"
subtitle: "Literate Data Exploration: Checking the ronaVote Regression Using Within-States Variation"

author: "Steven Bhardwaj"
date: "Posted 2021-08-13, Updated 2021-08-13"
output: 
  html_document:
    theme:
      bootswatch: darkly
    code_folding: hide
---

<style type="text/css">

h1.title {
  padding-top: 2em;
  text-align: center;
}
h3.subtitle {
  text-align: center;
}
h4.author {
  font-size: 1em;
  text-align: center;
}
h4.date {
  font-size: 1em;
  text-align: center;
}

#homeLink {
  font-size: 1.5em;
  color: #CCCCCC !important;
  opacity: 1;
  width: 5em;
	height: 1.2em;
  position: absolute;
  top: 3%;
  left: 50%;
  transform: translate(-50%, 0%);
}
hr{
  height: 1px;
  background-color: #FFF;
  border: none;
}
.center_plots {
  font-size: 1em;
  text-align: center;
  margin: auto;
  width: 70%;
}
.width_auto {
  margin: auto;
  width: 100%;
}
</style>


<div id="homeLink"><a href="../index.html"><p align="center">
bhrdwj.net
<br>
潘
</p></a></div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# class.source="fold-show"
```

## Summary
This is a literate-code walkthrough of the data exploration process, leaving the resulting charts at the end.

<hr> #### Libraries and Startup
```{r warning=FALSE , message=FALSE}
library(tidyverse)
library(broom)
library(rmarkdown)
library(ggdark)       # for dark_theme_gray()
library(stats)        # for weighted.mean()
library(ggiraph)      # for tooltip highlighting
library(directlabels) # for plot labeling, on within between graph

rm(list=ls())
load("./temp_workspace/rona.RData")

```

<hr> #### Do the states have enough counties to run "within" regressions?

```{r warning=FALSE , message=FALSE}
# group_by states
# regress, tidy, plot all-on-one graph (geom_line instead of geom_point)
summaries_ByState <- ronaPer1000 %>% 
  group_by(state) %>%
  summarize(mean_margin = weighted.mean(DJT_Margin, (REPUBLICAN+DEMOCRAT)),
            county_counts = n(),
            state_pop = sum(pop2019)
            )

plot_byState <- 
  ggplot(summaries_ByState, 
       aes(x = county_counts, y = mean_margin, size=state_pop)
       ) +
  labs(title="Vote-Margin and Number-Of-Counties, by State", 
       subtitle="3,104 out of 3,143 Represented in available data."
      ) +
  geom_point_interactive(aes(tooltip = state), alpha=0.5) +
  dark_theme_gray()

# rm(summaries_ByState)

girafe(code = print(plot_byState), width_svg = 6, height_svg = 4,
       options = list(opts_sizing(width = .75)))

```

<hr> #### Most states have enough counties, but some have too few. 
- I will combine states with fewer than 25 counties with other contiguous states into a few "regions".

```{r warning=FALSE, message=FALSE}

fips_stateOrRegion <- ronaPer1000 %>%
  mutate(newEngland = state %in% c("NH", "VT", "RI", "MA", "ME", "CT")) %>%
  mutate(AZ_NV = state %in% c("AZ", "NV")) %>%
  mutate(WY_UT = state %in% c("WY", "UT")) %>%
  mutate(DE_MD_NJ = state %in% c("DE", "MD", "NJ")) %>%
  mutate(stateOrRegion = ifelse(newEngland, "New_England", state)) %>%
  mutate(stateOrRegion = ifelse(AZ_NV, "AZ_NV", stateOrRegion)) %>%
  mutate(stateOrRegion = ifelse(WY_UT, "WY_UT", stateOrRegion)) %>%
  mutate(stateOrRegion = ifelse(DE_MD_NJ, "DE_MD_NJ", stateOrRegion)) %>%
  select(c(stateOrRegion, fips))

stateOrRegions_only <- fips_stateOrRegion$stateOrRegion %>% 
  unique() 


summaries_ByStateOrRegion <- ronaPer1000 %>% 
  left_join(fips_stateOrRegion, by="fips") %>%
  filter(! state %in% c("AK", "HI")) %>%
  select(-c(state)) %>%
  group_by(stateOrRegion) %>%
  summarize(mean_margin = weighted.mean(DJT_Margin, pop2019),
            county_counts = n(),
            state_pop = sum(pop2019),
            )

plot_byStateOrRegion <- 
  ggplot(summaries_ByStateOrRegion, 
       aes(x = county_counts, y = mean_margin, size=state_pop)
       ) +
  xlim(0, NA) +
  labs(title="Vote-Margin and Number-Of-Counties, by StateOrRegion", 
    caption=""
    ) +
  geom_point_interactive(aes(tooltip = stateOrRegion), alpha=0.5) +
  dark_theme_gray()

# rm(summaries_ByStateOrRegion)

girafe(code = print(plot_byStateOrRegion), width_svg = 8, height_svg = 4,
        options = list(opts_sizing(width = .75)))
```

<hr> #### Setup data for "within" and "between" regressions
```{r eval=TRUE, warning=FALSE, message=FALSE}
ronaPer1000_withStateOrRegion <- ronaPer1000 %>%
  left_join(fips_stateOrRegion, by="fips") %>%
  arrange(desc(stateOrRegion)) %>%
  filter(! state %in% c("AK", "HI")) %>%
  select(-c(state))

```
<hr> #### Declare a Function for time-series regressions
```{r warning=FALSE, message=FALSE}
regressEachWeekCol <- function(df, sR){   # sR stands for stateOrRegion
  lmOut_raw <- map(select(df, starts_with("T", ignore.case=FALSE)), 
              function(cas_per_1000) {
                lm(cas_per_1000 ~ DJT_Margin, df, weights = pop2019)
                })
  
  lmOut_stats_list <- map(lmOut_raw, 
               function(an_lm) {
                   c( 
                      coeff = tidy(an_lm)$estimate[2], 
                      y_intercept = tidy(an_lm)$estimate[1], 
                      pVal_coeff = tidy(an_lm)$p.value[2], 
                      pVal_intercept = tidy(an_lm)$p.value[1], 
                      r2 = glance(an_lm)$r.squared
                   )})
  
  lmOut_stats_df <- as.data.frame(do.call(rbind, lmOut_stats_list)) %>%
    mutate(stateRegion = sR) %>%
    rownames_to_column(var = "week_DateT")

  return(lmOut_stats_df)
}

```

<hr> #### Run state-by-state regressions
``` {r eval=FALSE, warning=FALSE, message=FALSE}

lmOut_stateOrRegion <- list()
for (sR in stateOrRegions_only[! stateOrRegions_only %in% c("HI", "AK")]) {
  lmOut_stateOrRegion[[sR]] <- ronaPer1000_withStateOrRegion %>% 
    filter(stateOrRegion == sR) %>%
    regressEachWeekCol(sR)
}

lmOut_betaCoeffs <- bind_rows(lmOut_stateOrRegion, .id = "column_label") %>%
  inner_join(weekDates, by = "week_DateT") %>%
  mutate(ronaVoteTrend = ifelse(pVal_coeff < .05 & coeff > 0, "Republican", 
                       ifelse(pVal_coeff < .05 & coeff < 0, "Democrat", NA))
         ) 


# SAVE TO lmOut_stateByState.RData
```

<hr> #### Plot state-by-state regressions
```{r warning=FALSE, message=FALSE, fig.height=8, fig.align='center'}

load("./temp_workspace/lmOut_stateByState.RData")

# ADD THE BETWEEN REGRESSIONS TO THIS CHART
# load("./temp_workspace/lmOut_stateByState.RData")

lmOut_betaCoeffs %>%
  ggplot(aes(x=weekDate, y=stateRegion, color=ronaVoteTrend)) +
  geom_point(size = 0.8) +
  scale_colour_manual(values = c(Republican = "#DD2222", Democrat = "#2222DD")) + 
  labs(title="Individual StateOrRegion Regression Coefficients Over Time",
       subtitle="Significant correlations highlighted and color-coded by sign.",
       caption=paste("State/Regions on weeks where (β>0) & (β=0 unlikely with p<0.05) have a red dot.\n",
                     "(β<0) & (β=0 unlikely with p<0.05) have a blue dot.\n",
                     "When (p>.05), then no dot is plotted."),
                     x="Date", y="StateOrRegion") +
  dark_theme_gray()
   
```

<hr> #### Run and plot coefficients "between" regression.  
- (Regression on stateOrRegions instead of on counties.)
```{r warning=FALSE, message=FALSE, fig.align='center'}
# rm(list=ls())
load("./temp_workspace/rona.RData")
load("./temp_workspace/fips_stateOrRegion.RData")

statesBetweenData <- ronaSectionsTot %>%
  left_join(fips_stateOrRegion) %>%
  filter(! state %in% c("AK", "HI")) %>%
  group_by(stateOrRegion) %>%
  summarize(REPUBLICAN = sum(REPUBLICAN),
            DEMOCRAT = sum(DEMOCRAT),
            DJT_Margin = (REPUBLICAN - DEMOCRAT) / (REPUBLICAN + DEMOCRAT),
            countyCounts_sR = n(),
            pop2019 = sum(pop2019),
            across(starts_with("T", ignore.case=FALSE), ~ sum(.)*1000/pop2019)
            )

betweenRegression_timeSeries <- regressEachWeekCol(statesBetweenData, 1) %>% 
  inner_join(weekDates)
  
ggplot(betweenRegression_timeSeries, aes(y=coeff, x=weekDate)) +
  geom_point() +
  labs(title='"Between" Regression Coefficients Time Series', 
       subtitle="subtitle", 
       caption="caption",
       x="Date",
       y="Coefficient ") +
  geom_hline(yintercept = 0, linetype = "solid", color = "white", size = .5) +
  geom_point(aes(color=coeff)) +
  scale_colour_gradient(low="#0000ff", high="#ff0000", guide=FALSE) + 
  dark_theme_gray()


```
<hr> #### Run and plot "within" (FE) regression coefficient as time series.
- (Inserting fixed-effect control regressors for each state or region, discarding the state/region effect coefficients.)

``` {r warning=FALSE, message=FALSE, fig.align='center'}
load("./temp_workspace/lmOut_stateByState.RData")

lmOutWithin_raw <- ronaPer1000_withStateOrRegion %>% {
  map(select(., starts_with("T", ignore.case=FALSE)), 
      function(cas_per_1000) {
        lm(cas_per_1000 ~ DJT_Margin + as.factor(stateOrRegion), data=., weights = pop2019)
        })
}

lmOutWithin_statsList <- map(lmOutWithin_raw, 
             function(an_lm) {
                 c( 
                    coeff = tidy(an_lm)$estimate[2], 
                    y_intercept = tidy(an_lm)$estimate[1], 
                    pVal_coeff = tidy(an_lm)$p.value[2], 
                    pVal_intercept = tidy(an_lm)$p.value[1], 
                    r2 = glance(an_lm)$r.squared
                 )})

withinRegression_timeSeries <- as.data.frame(do.call(rbind, lmOutWithin_statsList)) %>%
  rownames_to_column(var = "week_DateT") %>%
  inner_join(weekDates)

ggplot(withinRegression_timeSeries, aes(y=coeff, x=weekDate)) +
  geom_point() +
  labs(title='"Within" Regression Coefficients Time Series', 
       subtitle="subtitle", 
       caption="caption",
       x="Date",
       y="Coefficient ") +
  geom_hline(yintercept = 0, linetype = "solid", color = "white", size = .5) +
  geom_point(aes(color=coeff)) +
  scale_colour_gradient(low="#0000ff", high="#ff0000", guide=FALSE) + 
  dark_theme_gray()

```

<hr> #### Combining the "between" and "within" coefficient plots
-  We see that the state-to-state ronaVote trend is higher-magnitude than the trend among counties within states.
-  But there is broad similarity in the trend and timing.
```{r, warning=FALSE, message=FALSE, fig.align='center'}
a <- betweenRegression_timeSeries %>%
  select(coeff, weekDate) %>%
  mutate(regression = "between")

b <- withinRegression_timeSeries %>%
  select(coeff, weekDate) %>%
  mutate(regression = "within")

within_between_timeSeries <- rbind(a, b) 

g <- ggplot(within_between_timeSeries, 
       aes(y=coeff, 
           x=weekDate, 
           color=coeff, 
           linetype=factor(regression))) +
  geom_point() +
  scale_colour_gradient(low="#0000ff", high="#ff0000", guide=FALSE) + 
  labs(title='"Within" and "Between" Regression Coefficients', 
       subtitle="Covid Cases Per 100,000 Population Per .01 Vote Margin", 
       caption="caption",
       x="Date",
       y="Coefficient ") +
  geom_hline(yintercept = 0, linetype = "solid", color = "white", size = .5) +
  dark_theme_gray() +
  geom_dl(aes(label = regression), method = list(dl.combine("first.points", "last.points"), dl.trans(x=x-3, y=y-.5), dl.move("within", hjust=0, vjust=-2.7)), cex = 0) 

g

```




<!-- # EXPORT TO lmOut_stateOrRegion.RData -->
<!-- # myVec <- ls()[!ls() %in% c("lmOut_betaCoeffs", -->
<!-- #                            "weekDates", -->
<!-- #                            "stateOrRegions_only", -->
<!-- #                            "ronaPer1000_withStateOrRegion")] -->
<!-- # rm(list = myVec) -->
<!-- # rm(myVec) -->
