---
title: "ronaVote Appendix: Within"
subtitle: "Literate Data Exploration: Checking the ronaVote Regression Using Within-States Variation"

author: "Steven Bhardwaj"
date: "Posted 2021-08-13, Updated 2021-08-13"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme:
      bootswatch: darkly
    code_folding: hide
---

<style type="text/css">

h1.title {
  padding-top: 2em;
  text-align: center;
}
h3.subtitle {
  text-align: center;
}
h4.author {
  font-size: 1em;
  text-align: center;
}
h4.date {
  font-size: 1em;
  text-align: center;
}

#homeLink {
  font-size: 1.5em;
  color: #CCCCCC !important;
  opacity: 1;
  width: 5em;
	height: 1.2em;
  position: absolute;
  top: 3%;
  left: 50%;
  transform: translate(-50%, 0%);
}
hr{
  height: 1px;
  background-color: #FFF;
  border: none;
}
.center_plots {
  font-size: 1em;
  text-align: center;
  margin: auto;
  width: 70%;
}
.width_auto {
  margin: auto;
  width: 100%;
}
</style>


<div id="homeLink"><a href="../index.html"><p align="center">
bhrdwj.net
<br>
潘
</p></a></div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# class.source="fold-show"
```

## Summary
This is a literate-code walkthrough of the data exploration process, leaving the resulting charts at the end.

#### Libraries and Startup
```{r warning=FALSE , message=FALSE}
library(tidyverse)
library(broom)
library(rmarkdown)
library(ggdark)       # for dark_theme_gray()
library(stats)        # for weighted.mean()
library(ggiraph)      # for tooltip highlighting

rm(list=ls())
load("./temp_workspace/rona.RData")
rm(ronaSectionsTot)

```

## Do the states have enough counties to run "within" regressions?

```{r warning=FALSE , message=FALSE}

ronaPer1000 %>% count()

# group_by states
# regress, tidy, plot all-on-one graph (geom_line instead of geom_point)
summaries_ByState <- ronaPer1000 %>% 
  group_by(state) %>%
  summarize(mean_margin = weighted.mean(DJT_Margin, pop2019),
            county_counts = n(),
            state_pop = sum(pop2019)
            )

gg_obj <- 
  ggplot(summaries_ByState, 
       aes(x = county_counts, y = mean_margin, size=state_pop)
       ) +
  labs(title="Vote-Margin and Number-Of-Counties, by StateOrRegion", 
    caption=""
    ) +
  geom_point_interactive(aes(tooltip = state), alpha=0.5) +
  dark_theme_gray()

girafe(code = print(gg_obj), width_svg = 8, height_svg = 4)

```

Most states have enough counties, but some have too few. I will combine states with fewer than 25 counties with other contiguous states into a few "regions".

```{r}

fips_stateOrRegion <- ronaPer1000 %>%
  mutate(newEngland = state %in% c("NH", "VT", "RI", "MA", "ME", "CT")) %>%
  mutate(AZ_NV = state %in% c("AZ", "NV")) %>%
  mutate(WY_UT = state %in% c("WY", "UT")) %>%
  mutate(DE_MD_NJ = state %in% c("DE", "MD", "NJ")) %>%
  mutate(stateOrRegion = ifelse(newEngland, "New_England", state)) %>%
  mutate(stateOrRegion = ifelse(AZ_NV, "AZ_NV", stateOrRegion)) %>%
  mutate(stateOrRegion = ifelse(WY_UT, "WY_UT", stateOrRegion)) %>%
  mutate(stateOrRegion = ifelse(DE_MD_NJ, "DE_MD_NJ", stateOrRegion)) %>%
  select(c(stateOrRegion, fips))

summaries_ByStateOrRegion <- ronaPer1000 %>% 
  left_join(fips_stateOrRegion, by="fips") %>%
  filter(! state %in% c("AK", "HI")) %>%
  select(-c(state)) %>%
  group_by(stateOrRegion) %>%
  summarize(mean_margin = weighted.mean(DJT_Margin, pop2019),
            county_counts = n(),
            state_pop = sum(pop2019),
            )

gg_obj2 <- 
  ggplot(summaries_ByStateOrRegion, 
       aes(x = county_counts, y = mean_margin, size=state_pop)
       ) +
  xlim(0, NA) +
  labs(title="Vote-Margin and Number-Of-Counties, by StateOrRegion", 
    caption=""
    ) +
  geom_point_interactive(aes(tooltip = stateOrRegion), alpha=0.5) +
  dark_theme_gray()

girafe(code = print(gg_obj2), width_svg = 8, height_svg = 4)
```

## Run "within" regressions

```{r eval=TRUE}
ronaPer1000_byStateOrRegion <- ronaPer1000 %>%
  left_join(fips_stateOrRegion, by="fips") %>%
  arrange(stateOrRegion) %>%
  select(-c(state))

stateOrRegions_only <- ronaPer1000_byStateOrRegion %>% 
  distinct()

regressEachWeekCol <- function(df){

    # df <- ronaPer1000_byStateOrRegion %>%
    # filter(stateOrRegion == "New_England")
  
  x <- map(select(df, starts_with("T", ignore.case=FALSE)), 
              function(yvar) {
                lm(yvar ~ DJT_Margin, df, weights = pop2019)
                })
  
  lm_outT <- map(x, 
               function(an_lm) {
                   c( coeff = tidy(an_lm)$estimate[2], 
                      y_intercept = tidy(an_lm)$estimate[1], 
                      r2_coeff = glance(an_lm)$r.squared,
                      pVal_coeff = glance(an_lm)$p.value )
               })
  rm(df, x)
  return(lm_outT)
}


x <- ronaPer1000_byStateOrRegion %>%
  nest_by(stateOrRegion) %>%
  mutate(eachWeekRegLists = list(map(.[[2]], regressEachWeekCol)))



```

```{r eval=FALSE}
lm_allT <-  map(select(ronaPer1000_byStateOrRegion, starts_with("T", ignore.case=FALSE)), 
              function(yvar) {
                lm(yvar ~ DJT_Margin, ronaPer1000_byStateOrRegion, weights = pop2019)
                })


listOfStatesOrRegions_listOfWeeklyRegressions <- ronaPer1000_byStateOrRegion %>%
  group_by(stateOrRegion) %>%
  nest() %>%
  mutate(model = map())
  



lm_allT <-  map(select(ronaPer1000_byStateOrRegion, starts_with("T", ignore.case=FALSE)), 
              function(yvar) {
                lm(yvar ~ DJT_Margin, ronaPer1000_byStateOrRegion, weights = pop2019)
                })

# Tidy regression coefficients and other statistical output.
lm_outT <- map(lm_allT, 
               function(an_lm) {
                   c( tidy(an_lm)$estimate[2], 
                      glance(an_lm)$r.squared,
                      glance(an_lm)$p.value )
               })

lm_dfT <- as.data.frame(do.call(rbind, lm_outT))
colnames(lm_dfT) <- c("BetaCoeff", "RSquared", "PValue")
weekDates_ <- column_to_rownames(weekDates, var="week_DateT")

```

```{r eval=FALSE}

# Join additional columns for analysis
timeSeriesResults <- lm_dfT %>% 
  rownames_to_column(var = "week_DateT") %>%
  left_join(weekDates, by = "week_DateT") %>%
  column_to_rownames("week_DateT") %>%
  filter(weekDate > as.Date("2020-04-12"))

# Reorder columns
timeSeriesResults <- 
  timeSeriesResults[c("BetaCoeff", "RSquared", "PValue", "weekDate")]

rm(lm_outT, lm_dfT) # lm_allT, 


```


```{r eval=FALSE}
ts_title <- "title"
ts_subtitle <- "subtitle"
ts_caption <- "caption"

ts_plot <- timeSeriesResults %>% 
    ggplot(aes(x=weekDate, 
               y=BetaCoeff)) +
        labs(title=ts_title, subtitle=ts_subtitle, caption=ts_caption) +
        xlab("Date") + ylab("Regression Coefficient") +
        scale_y_continuous(limits=c(-10, 20)) +
        geom_hline(yintercept = 0, linetype = "solid", color = "white", size = .5) +
        geom_point(aes(color=BetaCoeff)) +
        scale_colour_gradient(low="#0000ff", high="#ff0000", guide=FALSE)

ts_plot + dark_theme_gray()


```
